### 2018-04-04 chapter6/ms2.tex
Redo Ch6. Start Intro

%UNLIKE IN A REGULAR TEX FILE, DON'T PUT ANY PREAMBLE MATERIAL HERE
\section*{Abstract}
We present preliminary studies into the development of efficient and effective strategies for the usage of LSST in optical follow-up of gravitational wave events. Specifically, we focus on the detection and characterization of kilonovae across a wide range of ejecta properties and composition. We show that the main LSST survey will achieve an average efficiency for kilonova detection of \apx3\%, with only a handful of sources characterized well enough for detailed study. In contrast, triggered target-of-opportunity follow up with a time investment of just 1 hour per event will be able to detect and characterize the brightest kilonova models, provided observations commence promptly after the GW trigger. This suggests that LSST has the opportunity to play an important role in the growth of multi-messenger astronomy by facilitating highly efficient kilonova searches.

@@

%and explore the detectability of kilonovae in triggered target-of-opportunity follow-up observations. We explore not only the ability of such observations to detect kilonovae, but also identify and characterize their behavior.

@@

\clearpage
\section{Simulated Observations}
\label{sec:ch6_obs}

%In this work, we consider the detectability and characterization of kilonovae in two unique observing situations. The first is the serendipitous detection of kilonovae in the LSST main survey independent of a GW trigger from LIGO. The second is triggered target-of-opportunity observations conducted by LSST in response to a GW trigger. We generate simulated observations for both scenarios as follows: 
%
%\subsection{Main Cadence Simulations With OpSim}
%\label{sec:ch6_opsim}

We produce simulated observations of kilonovae in the LSST main cadence using the LSST Operations Simulator \citep[OpSim,][]{OpSim1,OpSim2}. This is a publicly available package designed to produce realistic simulations of LSST scheduling and imaging over the ten year duration of the survey. These simulations provide realistic information about observations including cadence and science program requirements, observing conditions due to telescope or environmental factors, and image characteristics. We use the most recent OpSim reference run ({\tt minion\_1016}) to produce a catalog of observations including times, sky position, photometric errors, and limiting magnitudes.

We inject our kilonova model into this database of observations. For each injection, we uniformly choose a random model from the five described in \cref{sec:ch6_models}. The chosen model is injected at a randomly selected sky position and time, chosen uniformly. The model is injected at a random  
distance over a volume defined by the maximal LIGO BNS detection radius $(\apx450$ Mpc) in distance bins of 50~Mpc. Since it is not guaranteed that LSST with be looking at the chosen position and time where the source is injected, we continually inject sources until we have 100 kilonovae in each distance bin. This method allows more robust statistics at smaller distances compared to a volume-weighted injection scheme and allows us to compute LSST all-sky efficiencies and determine the total number of expected transients detected during the survey (see \cref{sec:ch6_opsim_results}).

%\subsection{Target of Opportunity Simulations With SNANA}
%\label{sec:ch6_snana}
%We parametrize our triggered target-of-opportunity (ToO) strategies based on the expected time investment per night and total number of nights spent observing. In this work we assume that the typical LIGO localization region is \apx10~deg$^2$ and can therefore be observed in just two LSST pointings, which are necessary to cover chip gaps in the CCD mosaic. We assume that each pointing consists of $grizY$ observations lasting 30~s each with per filter overheads of 90~s. This puts the total time investment at 10 minutes per pointing, or 20 minutes per night. We assume 7 nights of observing for a total telescope time investment of \apx2.5~hours. To test the effect of rapid cadence observations, we include a second epoch on the first night, taken 3 hours after the initial epoch.
%
%We simulate these ToO observations using the SNANA public software package \citep{SNANA}. SNANA is a supernova analysis and simulation package designed to produce realistic supernovae light curves for the optimization of time-domain surveys and extraction of cosmological parameters. SNANA generates these light curves using a model transient SED, realistic filter transmission functions, and a database of observation dates and conditions. We run SNANA using our kilonova model SEDs and LSST filter transmission functions, and generate the database of observing conditions by pulling key information from the OpSim simulations (e.g., sky noise, PSF, observed zero point, see \cref{sec:ch6_opsim}). We then mutate the observing dates to match those required for our ToO cadence described above.
%
%We inject kilonova following the same procedure outlined in \cref{sec:ch6_opsim} with several key differences. First, we do not inject sources at randon sky locations, but instead at the location LSST is observing in the SNANA database. Second, in order to probe a realistic range of response times for the ToO trigger, we inject kilonova at a uniformly sampled time from 3--24 hours prior to the first epoch. Lastly, SNANA randomly selects an injection distance using a volume-weighted scheme to generate a realistic distribution. We therefore, inject 10,000 kilonovae across the entire volume ($D_L \apx450$~Mpc) to ensure that we have enough nearby sources. Since the distance injection in SNANA is based on redshift, a handful of sources end up with a distance greater than 450~Mpc. To maintain consistency with our OpSim volume, we truncate these sources leaving a final sample of 9171 light curves for our analysis in \cref{sec:ch6_snana_results}.

@@

%In the context of targeted ToO observations, we are not only concerned with these issues of detection and characterization, but also the impact of program design (e.g., telescope time investment) on the science returns.

@@

Applying these criteria across all distances and all models we find that out of 3653 injected kilonova we detect 123 kilonovae giving a total efficiency of $3\%$. If we exclude the 2553 kilonovae that were injected at times and sky locations not observed by LSST then this efficiency increases to $11\%$. Furthermore, of the 123 kilonovae detected in our survey we able to obtain information about the rise for 62 events and color information for 99 events. Lastly, we are able to obtain both color and rise information for 52 events. These numbers are shown as a function of distance in \Cref{fig:ch6_eff_hist} and as a function of kilonova model in \cref{tab:ch6_eff}.

\begin{deluxetable}{ccccccc}
\singlespace
\tabletypesize{\footnotesize}
\tablecolumns{12}
\tablewidth{0pt}
\tablecaption{LSST Main Survey Efficiencies
	          \label{tab:ch6_eff}}
\tablehead{
\colhead{Model} &
\colhead{Name} &
\colhead{$N$} &
\colhead{$N_{\rm det}$} &
\colhead{$N_{\rm rise}$} &
\colhead{$N_{\rm color}$} &
\colhead{$N_{\rm both}$}
}
\startdata
1 &  GW170817 & 172 & 29 & 16 & 24 & 13 \\
2 & {\tt h\_blue\_h\_red} & 167 & 47 & 21 & 37 & 18 \\ 
3 & {\tt h\_blue\_l\_red} & 165 & 24 & 14 & 16 & 10 \\
4 & {\tt l\_blue\_h\_red} & 212 & 0 & 0 & 0 & 0 \\
5 & {\tt l\_blue\_l\_red} & 181 & 0 & 0 & 0 & 0
\enddata
\tablecomments{Detection efficiencies for each kilonova model. $N$ denotes the total number of models injected. $N_{\rm det}$ denotes the total number of kilonovae detected. $N_{\rm rise}$ denotes the total number of kilonovae for which brightening was detected. $N_{\rm color}$ denotes the total number of kilonovae for which color information was obtained. Lastly, $N_{\rm both}$ denotes the total number of sources for which both brightening and color information is available.}
\end{deluxetable}

%\begin{figure}[!t]
%\begin{center}
%\hspace*{-0.1in}
%\scalebox{1.}
%{\includegraphics[width=0.75\textwidth]{./figs/chapter6/f1.pdf}}
%\caption{\singlespace Histograms showing the number of detections as a function of model and distance, for those models that are detected. We note that as a result of the cadence of the LSST main survey we are not always able to obtain crucial information about the rise and color.}
%\label{fig:ch6_eff_hist}
%\end{center}
%\end{figure}

\clearpage
Looking at \Cref{fig:ch6_eff_hist} and \cref{tab:ch6_eff}, we first note that only the three brightest models (GW170817, {\tt h\_blue\_h\_red}, and {\tt h\_blue\_l\_red}) are detected. The model {\tt l\_blue\_h\_red}, is not detected despite having the same {\it total} ejecta mass as the model {\tt h\_blue\_l\_red}. This is because all the mass in the {\tt l\_blue\_h\_red} is contained in the high-opacity ``red" component and is therefore less luminous. The model {\tt l\_red\_l\_blue} is simply too faint to detect due to it's extremely low total ejecta mass.  Inspecting \Cref{fig:ch6_eff_hist} for trends that depend on distance, we note that the models for GW170817 and {\tt h\_blue\_h\_red} are preferentially detected at small distances ($D_L \lesssim 100$~Mpc). There is a similar trend for detecting brightening in these models, however there is no strong dependence on distance and the availability of color information. The model {\tt h\_blue\_l\_red} shows no strong dependence on distance for any criteria.

We also investigate the timing of our detections by defining two crucial times. The value $t_{\rm first}$ is the time, in days, from explosion to the first observation with S/N $> 5$, while $t_{\rm last}$ is the time from explosion to the last observation with S/N $> 5$. The number of events as a function of these two values are shown in \Cref{fig:ch6_t_hist}. Looking at the distribution with respect to $t_{\rm first}$, we find a slight preference for detections at early times particularly for the model of GW170817.  We also find that the latest value of $t_{\rm first}$ is \apx$5-6$~days for all three models. Looking at the distribution with respect to $t_{\rm last}$, we first find that in general the last observation is obtained within a week of explosion. The only exception is GW170817, where a handful of events are detected out to two weeks post explosion. We also note that 4 events have a value of $t_{\rm last}$ of $\lesssim 1$~day. These events appeared in high cadence regions of the survey, resulting in 4-5 observations being obtained in a single night. The sky positions of these events do not correspond to the LSST deep drilling fields, but rather a spurious increases in observing cadence.

%\begin{figure}[!t]
%\begin{center}
%\hspace*{-0.1in}
%\scalebox{1.}
%{\includegraphics[width=0.725\textwidth]{./figs/chapter6/f2.pdf}}
%\caption{\singlespace Histogram of events as a function of time to first or last detection. As with \Cref{fig:ch6_eff_hist}, we only plot those models that are detected. The LSST main cadence is only able to track a limited number of events out to longer times than \apx~a few days. This combined with the poor rise and color constraints shown in \Cref{fig:ch6_eff_hist} means that kilonovae detected in the LSST main survey will be difficult to detect and characterize.}
%\label{fig:ch6_t_hist}
%\end{center}
%\end{figure}

Lastly, we are concerned with the overall efficiency of our search. This is shown as a function of distance and model in \Cref{fig:ch6_dist_eff}. As expected, the survey achieves the highest efficiency for nearby sources $(D_L \apx50$~Mpc). If we only consider injected sources that were observed by LSST, we compute a model averaged efficiency of \apx26\%. If we exclude models that are never detected, this efficiency rises to \apx48\%. This efficiency drops sharply to \apx15\% at a distance of $D_L \apx250$~Mpc, beyond which the efficiency shows no strong dependence on distance. The efficiency averaged over the entire volume considered is \apx11\% for all models, and \apx17\% when considering only detected models.

%\begin{figure}[!t]
%\begin{center}
%\hspace*{-0.1in}
%\scalebox{1.}
%{\includegraphics[width=0.9\textwidth]{./figs/chapter6/f3.pdf}}
%\caption{\singlespace Detection efficiency as a function of distance for each of the detected models. As expected, the efficiency is high (\apx45\%) for nearby sources ($D_L \apx50$~Mpc), but rapidly tapers off to \apx15\% at the edge of the LIGO BNS detection range $(D_L \gtrsim 400$~Mpc).}
%\label{fig:ch6_dist_eff}
%\end{center}
%\end{figure}

We now wish to compute the total number of kilonovae we expect to detect in our search volume over the course of the LSST main survey. We compute this number using the following expression:
\begin{equation}
N_{\rm tot} = 4 \pi \mathcal{R} \int \epsilon (z) \frac{dV}{dz} dz,
\end{equation}

\noindent where $\mathcal{R}$ is the expected rate of kilonovae, $\epsilon (z)$ is the efficiency as a function of distance, and $dV/dz$ is the differential comoving volume \citep[see e.g.,][]{Hogg1999}. The factor of $4\pi$ accounts for the fact that $dV/dz$ has units of Mpc$^{3}$ sr$^{-1}$ and indicates we are considering the number of kilonovae across the entire sky. We assume the rate of kilonovae is the rate of BNS mergers derived by LIGO during the second observing run \citep[$\mathcal{R} \apx1500$~Mpc$^{-3}$~yr$^{-1}$,][]{LIGOGW170817}. Here we compute the efficiency, $\epsilon (z)$, using the total number of kilonovae injected in each distance bin. We find a total expected number of $N_{\rm tot} \approx 8.8$ kilonovae per year across the entire ten year survey. This number is comparable to that found by \citet{Scolnic+18}. If we compute $\epsilon (z)$ by only considering sources for which there is both rise and color information, then we find $N_{\rm tot} \approx 3.6$ kilonovae per year.

The key point emerging from this study is that despite the large \'{e}tendue of the LSST survey, it will be relatively inefficient at finding kilonova compared to other types of transient events. More importantly, the small number of kilonovae that will have sufficient light curve data for more detailed studies suggests that the science returns for each event will be low. This low efficiency also suggests that the likelihood of a serendipitous joint detection between LIGO and LSST is vanishingly small. Even in the fortuitous situation of a joint detection, if it takes several days to detect and identify the kilonovae, this will severely limit our ability to perform multi-wavelength imaging and spectroscopy. These comprehensive multi-wavelength follow-up programs were found to be crucial for maximizing the science returns from GW170817.

\subsection{Results from SNANA}
\label{sec:ch6_snana_results}
We analyze the sample of light curves from our SNANA simulations using the same criteria outlined in \cref{sec:ch6_opsim_results}, with one key change: we now require that brightening be detected between observations taken in the same filter. This is necessary as our broadband imaging on a single night will sample the kilonova SED well enough that erroneous "brightening" will be detected between filters. We apply these criteria to our sample of light curves finding that as with the OpSim simulations, we are only able to detect the three brightest kilonovae models. This is expected given that our ToO observations still use the same exposure time as OpSim (30~s), so even with an improved cadence and broadband coverage the models {\tt l\_blue\_h\_red} and {\tt l\_blue\_h\_red} are simply not bright enough to detect without deeper exposures.

The detection efficiency as a function of distance is shown in \Cref{fig:ch6_snana_dist_eff}. We first note that the efficiency is flat as a function of distance with a model-averaged efficiency of \apx20\%. It is important to note however, that this number is computed using all injected sources. If we consider only those models for which there is at least one detection, then the efficiency is 100\% at all distances. Furthermore, as a result of the nightly broadband imaging we are able to obtain color information for all detected kilonovae. However, we find that our ability to detect brightening in the light curve is poor, with an average efficiency of \apx10\%. This efficiency shows no strong dependence on distance, suggesting that other factors such as the promptness of observations are important.

%\begin{figure}[!t]
%\begin{center}
%\hspace*{-0.1in}
%\scalebox{1.}
%{\includegraphics[width=0.9\textwidth]{./figs/chapter6/f5.pdf}}
%\caption{\singlespace Efficiency of our ToO searches as a function of distance. The numbers plotted here are relative to the total number of injected sources (i.e., including models not detected). The {\it per model} detection efficiency is 100\% for the bright KN models plotted. The efficiency curve is flat as a function of distance indicating that ToO efforts with LSST can be effective out to the maximal LIGO BNS detection distance.}
%\label{fig:ch6_snana_dist_eff}
%\end{center}
%\end{figure}

We investigate this by computing our efficiency as a function of time, in hours, between the time of explosion and the time of our first observation. We first note that the overall detection efficiency is not affected by the start time of our observations. We again achieve a model-averaged efficiency of \apx20\% when considering all injected sources, and an efficiency of 100\% when only considering the models for which we have detections. However, we do find that our ability to detect brightening in the light curve is strongly impacting by the start time of observations. This can be seen in \Cref{fig:ch6_snana_t_eff} which shows the fraction of detected sources that also exhibit brightening in at least one filter. For observations that start promptly (+3~hours), then brightening can be detected in 60\% of sources. This however, plummets quickly to \apx5\% at +9 hours, and 0\% at later times. This highlights the fact that rapid response must be a priority for any triggered ToO program to be successful.

%\begin{figure}[!t]
%\begin{center}
%\hspace*{-0.1in}
%\scalebox{1.}
%{\includegraphics[width=0.9\textwidth]{./figs/chapter6/f6.pdf}}
%\caption{\singlespace Efficiency of our ToO searches to detect brightening in the kilonova light curve in at least one filter as a function of starting time. The efficiency here is computed relative to the number of detected sources in a given time bin $(\Delta_t \apx3$~hrs). There is a sharp drop off in efficiency after just a few hours. This efficiency drops to 0 under all observing scenarios when just a single epoch is obtained on the first night.}
%\label{fig:ch6_snana_t_eff}
%\end{center}
%\end{figure}

Lastly, we consider the effect of altering our ToO cadence and telescope time investment in our computed efficiencies. First, we investigate the effect of removing the second epoch from the first night of observations. This has no impact on our general ability to detect kilonovae, resulting in identical efficiency calculations. There is however, a strong impact on our ability to detect the kilonova rise. When observations are conducted without the second epoch on the first night, the efficiency for detecting brightening drops to 0\% for all models, at all distances and all start times. The second rapid epoch is therefore crucial for characterizing the early light curve behavior. Second, we investigate the effect of truncating the total time spent observing to 3 or 5 days per event. This is accomplished by taking the simulated light curves, removing epochs after the truncation time, and recomputing all of our criteria. We find that truncating our light curves does not alter our efficiency calculations. This suggests that just 20 minutes a night for 3 nights (e.g., one hour total telescope time) is sufficient to detect the brightest kilonovae out to the maximal LIGO BNS detection range. Candidates identified in this manner could be passed off to other facilities for targeted follow-up (e.g., deep imaging and spectroscopy) allowing LSST to quickly return to normal science operations.

\section{Discussion and Conclusion}
\label{sec:ch6_conc}
We have presented detailed simulated observations of kilonovae using LSST in two distinct contexts. First, we investigated the ability of LSST to both detect and characterize kilonovae over the course of the primary ten year survey. Second, we investigate the effectiveness of a triggered target-of-opportunity campaign conducted in response to a gravitational wave event trigger from the Advanced LIGO/Virgo interferometers. A key point that emerges from both studies is that it may be difficult to detect kilonova across the entire range of possible ejecta parameters. Models that have ejecta masses similar to or greater than those associated with GW170817 can be detected by LSST. However, models with less ejecta mass quickly become too faint to observe effectively. We further find:

\begin{enumerate}
\item We find that the LSST main survey has an average efficiency for kilonova detection of \apx3\%. This efficiency is strongly a function of distance, peaking for nearby sources $(D_L \lesssim 100$~Mpc). Using this efficiency, we find that LSST will detect \apx8.8 kilonovae per year in the main survey over ten years. However, if we consider our efficiency for sources with enough light curve information to perform detailed studies, this number drops to \apx3.6 kilonovae per year.
\item We find that triggered ToO observations with LSST can be highly effective at detecting and characterizing kilonovae compared to the main survey cadence. ToO campaigns involving just an hour of telescope time are able to achieve a model-averaged efficiency of \apx45\%. This efficiency is 100\% for the brightest kilonova models. A key concern is prompt scheduling of observations in response to a trigger, as crucial early time behavior is difficult to capture for observations starting later than \apx3-6 hours post-explosion.
\end{enumerate}

Most importantly, we have shown that the cadence of the LSST main survey is poorly matched to kilonovae timescales, leading to just a handful of detections per year. However, targeted ToO campaigns with a modest investment in telescope time (\apx1~hour) can detect the brightest kilonovae models across the entire Advanced LIGO/Virgin volume. An increase in this time investment by a factor of 2 to account for deeper exposures will allow LSST to probe the faintest kilonovae models at a level that is simply not possible with other ground-based survey instruments. Therefore, it is clear that LSST can and should play a vital and important role in the future of joint GW-EM astronomy. 
%NO BIB INFO

>>>>

\section*{Abstract}
We present preliminary studies into the development of efficient and effective strategies for the usage of LSST in optical follow-up of gravitational wave events. Specifically, we focus on the detection and characterization of kilonovae across a wide range of ejecta properties and composition. We show that the main LSST survey will achieve an average efficiency for kilonova detection of \apx3\%, with only a handful of sources characterized well enough for detailed study. In contrast, triggered target-of-opportunity follow up with a minimal time investment of a few hours per night will be able to detect a wider range of kilonova models, provided observations commence promptly after the GW trigger. This suggests that LSST has the opportunity to play an important role in the growth of multi-messenger astronomy by facilitating highly efficient kilonova searches.

@@

@@

\clearpage
\section{Simulated Observations with OpSim}
\label{sec:ch6_obs}
We produce simulated observations of kilonovae in the LSST main cadence using the LSST Operations Simulator \citep[OpSim,][]{OpSim1,OpSim2}. This is a publicly available package designed to produce realistic simulations of LSST scheduling and imaging over the ten year duration of the survey. These simulations provide realistic information about observations including cadence and science program requirements, observing conditions due to telescope or environmental factors, and image characteristics. We use the most recent OpSim reference run ({\tt minion\_1016}) to produce a catalog of observations including times, sky position, photometric errors, and limiting magnitudes.

We inject our kilonova model into this database of observations. For each injection, we uniformly choose a random model from the five described in \cref{sec:ch6_models}. The chosen model is injected at a randomly selected sky position and time, chosen uniformly. The model is injected at a random  
distance over a volume defined by the maximal LIGO BNS detection radius $(\apx450$ Mpc) in distance bins of 50~Mpc. Since it is not guaranteed that LSST with be looking at the chosen position and time where the source is injected, we continually inject sources until we have 100 kilonovae in each distance bin. This method allows more robust statistics at smaller distances compared to a volume-weighted injection scheme and allows us to compute LSST efficiencies and determine the total number of expected transients detected during the survey (see \cref{sec:ch6_opsim_results}).

@@

@@

\clearpage
Applying these criteria across all distances and all models we find that out of 3653 injected kilonova we detect 123 kilonovae giving a raw efficiency of $3\%$. Of the 123 kilonovae detected in the survey, we able to obtain information about the rise for 62 events and color information for 99 events. We are able to obtain both color and rise information for 52 events. These are simply the raw numbers from our simulations. If is more physically interesting to compute the total number of kilonovae we expect to detect in our search volume over the course of the LSST main survey. We compute this value using the following expression:

\begin{equation}
N_{\rm tot} = 4 \pi \mathcal{R} \int \epsilon (z) \frac{dV}{dz} dz,
\end{equation}

\noindent where $\mathcal{R}$ is the expected rate of kilonovae, $\epsilon (z)$ is the efficiency as a function of distance, and $dV/dz$ is the differential comoving volume \citep[see e.g.,][]{Hogg1999}. The factor of $4\pi$ accounts for the fact that $dV/dz$ has units of Mpc$^{3}$ sr$^{-1}$ and indicates we are considering the number of kilonovae across the entire sky which is then corrected by the efficiency. We assume the rate of kilonovae is the rate of BNS mergers derived by LIGO during the second observing run \citep[$\mathcal{R} \apx1500$~Mpc$^{-3}$~yr$^{-1}$,][]{LIGOGW170817}. Here we compute the efficiency, $\epsilon (z)$, using the total number of kilonovae injected in each distance bin. We find a total expected number of $N_{\rm tot} \approx 8.8$ kilonovae per year across the entire ten year survey. This number is comparable to that found by \citet{Scolnic+18}. We compute $\epsilon (z)$ by only considering sources that satisfy our criteria beyond a simple detection. This will give the number of kilonovae expected per year which contain more detailed light curve information. These numbers are shown as a function of distance in \Cref{fig:ch6_eff_dist}.

\begin{figure}[!t]
\begin{center}
\hspace*{-0.1in}
\scalebox{1.}
{\includegraphics[width=0.95\textwidth]{./figs/chapter6/f1a.pdf}}
\caption{\singlespace Expected number of kilonova per year per distance bin, computed as described in \cref{sec:ch6_opsim_results}. The dip in expected numbers at 300~Mpc is simply the result of stochasticity in the simulation.}
\label{fig:ch6_eff_dist}
\end{center}
\end{figure}

\clearpage
Looking at \Cref{fig:ch6_eff_dist}, we first note that only the three brightest models (GW170817, {\tt h\_blue\_h\_red}, and {\tt h\_blue\_l\_red}) are detected. The model {\tt l\_blue\_h\_red}, is not detected despite having the same {\it total} ejecta mass as the model {\tt h\_blue\_l\_red}. This is because all the mass in the {\tt l\_blue\_h\_red} is contained in the high-opacity ``red" component and is therefore less luminous. The model {\tt l\_red\_l\_blue} is simply too faint to detect due to it's extremely low total ejecta mass. Inspecting \Cref{fig:ch6_eff_dist} for trends that depend on distance, we first note that there is a dip in the expected number at 300~Mpc. This feature can be attributed to stochasticity in the simulations. There is also a peak in the expected numbers around 350--400~Mpc. This represents the optimal position where there is a large volume probed, but sources are still nearby enough to be detected. Lastly, as expected, requiring stricter cuts on light curve behavior results in fewer detections.

The final point worth considering here is that this analysis has relied on an extremely minimal set of detection and characterization criteria. The light curves of kilonovae that satisfy these criteria will be sparsely sampled in both time and broadband filter coverage and we will not be able to make definitive statements about key properties such as color and timescale. As a result, it will in practice be difficult if not impossible to claim these as definitive kilonovae detections. Consequently, while a handful of kilonovae may appear in the LSST survey per year the overall science returns per event will be low.

\clearpage
\subsection{Benefits of Target of Opportunity Observations}
\label{sec:ch6_too_benefits}
The key point emerging from \cref{sec:ch6_opsim_results} is that despite the large \'{e}tendue of the LSST survey, it will be relatively inefficient at finding kilonova compared to other types of transient events. This low efficiency also suggests that the likelihood of a serendipitous joint detection between LIGO and LSST is vanishingly small. Even in the fortuitous situation of a joint detection, if it takes several days to detect and identify the kilonovae, this will severely limit our ability to perform multi-wavelength imaging and spectroscopy. These comprehensive multi-wavelength follow-up programs were found to be crucial for maximizing the science returns from GW170817.

The most obvious and effective way to combat this issue is by engaging in triggered target-of-opportunity observations. Consider such a campaign that invests no more than an hour or two per night for a few days to obtain deep images of the localization region with several broadband filters. These observations will be able to achieve depths that are 1.5--2 magnitudes deeper than the LSST main survey and consequently probe larger regions of the kilonovae ejecta parameter space (e.g., models such as {\tt l\_blue\_h\_red} and {\tt l\_red\_l\_blue}). These observations should commence promptly after the GW trigger is received to ensure that the crucial early-time light curve behavior is adequately captured. Furthermore, obtaining broadband imaging covering multiple filters allows accurate characterization of the rapidly evolving kilonova colors in a way that can not be achieved with sparse observations. Once a clear kilonova candidate (or handful of candidates) has been identified, further follow-up photometric and spectroscopic observations can be conducted using other telescopes allowing LSST to quickly return to main science operations.

\section{Discussion and Conclusion}
\label{sec:ch6_conc}
We have presented detailed simulated observations of kilonovae using LSST while considering two distinct contexts. First, we investigated the ability of LSST to both detect and characterize kilonovae over the course of the primary ten year survey. Second, we discuss the effectiveness of a triggered target-of-opportunity campaign conducted in response to a gravitational wave event trigger from the Advanced LIGO/Virgo interferometers. A key point that emerges from both considerations is that it may be difficult to detect kilonova in the main survey across the entire range of possible ejecta parameters. Models that have ejecta masses similar to or greater than those associated with GW170817 can be detected by LSST. However, models with lower ejecta mass quickly become too faint to observe effectively. We also show that the cadence of the LSST main survey is poorly matched to kilonovae timescales, leading to just a handful of detections per year.

We instead argue that targeted ToO campaigns with a modest investment in telescope time (\apx~a~few~hours) can detect a wider range of kilonovae models across the entire Advanced LIGO/Virgo volume. The resulting kilonovae light curves from such a campaign will be far better sampled in both time and wavelength compared to those detected in the main survey, allowing detailed studies of the kilonovae behavior and properties. From this work, a clear picture emerges that LSST can and should play a vital and important role in the future of joint GW-EM astronomy.

### 2018-04-04 introduction/intro.tex
Redo Ch6. Start Intro

%No preamble material
Grandiose intro paragraph about joint GW-EM astronomy

Top level description of thesis

Outline of this chapter.

\section{Gravitational Waves}
\label{sec:intro_gw}
Brief history of gravitational waves: GR, Einstein's Paper, Basic idea of what they are.

Very brief mathematical introduction

Argument that they must be quadrupole and resulting polarization

Something about translating to radiation and $1/r$ relation

Intuition about strain scales

@@

1-2 paragraphs on CMB, background, SMBH, etc.

Relevant to this thesis is compact object binary mergers

@@

The GRB.

@@

Radio and X-Ray stuff.

Slightly more detailed outline of thesis compared to preamble

>>>>

%No preamble material
The direct detection of gravitational waves from distant astrophysical sources has opened up and new and exciting view of the cosmos, not unlike the first detections of celestial objects in radio or x-rays. Gravitational wave observations allow new probes of Einstein's General Theory of Relativity in regimes not previously accessible and illuminate the existence of astrophysical objects that were previously invisible to purely electromagnetic observations. The true power of gravitational wave detections become apparent when they are paired with information from traditional electromagnetic telescopes. This emerging field of joint gravitational wave-electromagnetic (GW-EM) astronomy presents an unparalleled opportunity to test our understanding of the Universe.

This thesis presents a series of studies, both observational and theoretical, into developing and executing observational programs that facilitate the detection of an electromagnetic counterpart coincident with a gravitational wave detection. These studies first focus on detailed investigations into the design of optical follow-up programs that overcome many of the hurdles expected to arise when searching for these elusive electromagnetic counterparts. More recently, this work has focused on actually conducting follow-up searches, including the first joint detection of gravitational waves and electromagnetic waves.

In this chapter, we present an introduction to the growing field of joint GW-EM astronomy. This includes a brief introduction to gravitational waves, their sources in the Universe, and how we detect them (\cref{sec:intro_gw}), an overview of the expected electromagnetic counterparts and how we conduct searches for them (\cref{sec:intro_joint}), and lastly a history of gravitational wave detections leading up to the first joint GW-EM discovery (\cref{sec:intro_history}).

\section{Gravitational Waves}
\label{sec:intro_gw}
The existence of gravitational waves (GW) was first investigated by Einstein shortly after he introduced the world to his general theory of relativity \citep{Einstein1916}. Gravitational waves are ripples in both space and time that propagate through the universe at the speed of light, carrying with them energy, momentum, and most importantly information about their progenitors. Mathematically, gravitational waves are most simply thought of as a small perturbation to a flat metric \citep[$g_{\mu\nu}$; ][]{Misner+73,Carroll2004}:
\begin{equation}
\label{eq:intro_metric}
g_{\mu\nu} = \eta_{\mu\nu} + h_{\mu\nu}, \quad ||h|| \ll 1
\end{equation}
\noindent where $\eta_{\mu\nu}$ is the flat Minkowski spacetime given by $\eta_{\mu\nu} = {\rm diag}(-1, 1,1,1)$ and $h_{\mu\nu}$ is called the spacetime {\em strain} and can be loosely thought of as the amplitude of the gravitational wave. This linearized approach leads to a solution representing gravitational plane waves \citep[see e.g.,][and references therein]{Misner+73,Carroll2004}.

While discussions of this plane wave solution and the detailed nature of gravitational waves are beyond the scope of this thesis, there are several important physical aspects of gravitational waves worth mentioning. Linear/Angular momentum -> quadrupole emission (equation).

@@

Relevant to this thesis is compact object binary mergers. 1-2 paragraphs

@@

GRB and multi-wavelength detections

@@
